Directionality and Direction of Innovation in the Swedish Forest-Based Bioeconomy Transition

### 2024-04-03 paper.qmd
update text

format:
    pdf:
        papersize: A4
        linestretch: 1.25
---

@@

The bioeconomy has gained relevance among policymakers, practitioners and researchers alike.
Broadly speaking, the bioeconomy refers to an economy in which raw materials for energy and production processes are provided from biological instead of fossil sources.
<!-- TODO: citation -->
Hopes of increased resource efficiency and renewability as well as promised economic growth especially in regions rich in biological resources abound [@el-chichakli2016PolicyFiveCornerstones].
However, a such a profound change to economic systems will not be possible without balancing significant tradeoffs between the needs and desires of different human and non-human stakeholders. @hogbom2021TrilemmaNordicBaltic summarize these tradeoffs as a trilemma between forests ability to provide sufficent wood to replace fossil based materials, carbon sequestration and storage, and the delivery of other eco-system services such as forest biodiversity.
Resolving these tradeoffs is challening, as goal conflicts between different development pathways make simulatneous progress on social, ecological and economic goals difficult [@bennich2021BiobasedEconomy2030].

@@

These pillars reflect the two most influential visions of the bioeconomy concept. Notions of a biotechnology and bio-resource oriented bioeconomy are widespread and dominate discourse at various levels [@dieken2021MultitudesBioeconomiesSystematic].

@@

In more recent years, visions of a bio-resource bioeconomy have taken over as the leading narrative. Unlike the biotechnology vision, this conceptualization does not only focus on economic growth and job creation, but attempts to include ecological sustainability challenges [@bugge2016WhatBioeconomyReview]. Biological resources, not technologies, form the basis of a new economy. In essence, there are two ways in which bio-resources are poised to replace fossil resources: as substitution, e.g., biofuels replacing fossil fuels, and as high value added products, e.g., cross-laminated timber used as an alternative to concrete in construction [@vivien2019HijackingBioeconomy]. To substitute fossil resources requires significant efforts in innovation. In contrast to the biotechnology vision, however, innovation is not confined to scientific advances from a narrow biochemical and pharmaceutical field. Instead, new and old actors are required to collaborate between and beyond established sectors [@bugge2016WhatBioeconomyReview]. Although sustainability plays a more important role in this vision of the bioeconomy, this vision is considered an example of weak sustainability [@damato2017GreenCircularBio; @vivien2019HijackingBioeconomy]. The reason is the focus of substitution and high biomass demands. The consumption of raw materials may even increase, not only for bio-based, but also for fossil-based resources [@asada2020BioeconomicTransitionProjecting].

An overview or more likely more elaborate description of @vivien2019HijackingBioeconomy and @bugge2016WhatBioeconomyReview here.
Tables as placeholders for now.

| | Biotechnology | Bioresource | Bioecology |
|---|---|---|---|
| Aim &   Objectives | Economic growth & job   creation | Economic growth &   sustainability | Sustainability, biodiversity,   conservation of ecosystems, avoiding soil degradation |
| Value   Creation | Application of biotechnology,   commercialisation of research & technology | Conversion and upgrading of   bio-resources (process oriented) | Development of integrated   production systems and high-quality products with territorial identity |
| Drivers   & mediators of innovation | R & D, patents, TTOs,   Research councils and funders (Science push, linear model) | Interdisciplinary, optimisation   of land use, include degraded land in the production of biofuels, use and   availability of bio-resources, waste management, engineering, science &   market (Interactive & networked production mode) | Identification of favourable   organic agro-ecological practices, ethics, risk, transdisciplinary   sustainability, ecological interactions, re-use & recycling of waste,   land use, (Circular and self-sustained production mode) |
: Table adapted from @bugge2016WhatBioeconomyReview [10] {#tbl-bugge}

|  | Type II | Type III | Type I |
|---|---|---|---|
|  | A science-based economy driven   by industrial biotechnology<br>     <br>     The cell is a factory | Biomass replaces fossil fuels   and mining to produce energy and materials<br>     <br>     Biorefining at the heart of ecological transition (multilevel perspective). | An ecological economy, that is   compatible with the biosphere<br>     <br>     Counter-expertise rather than concrete technical solutions Criticism from   social groups who remain at the margins of decisionmaking centers |
: Table adapted from @vivien2019HijackingBioeconomy [194] {#tbl-vivien}

@@

@@

### Directionality

@@

<!-- Key paper cited in every paper on directionoality
@weber2012LegitimizingResearchTechnology -->

@@

Based on the analyses of competing bioeconomy visions in @bugge2016WhatBioeconomyReview and @vivien2019HijackingBioeconomy, I develop and apply a matrix of characterstics for each vision.
<!-- TODO: Matrix That matrix is presented in @tbl- -->
Reading the source texts of innovations identified from SWINNO, I classify an innovation as being *aligned* with a vision if it fulfills at least one criteria in the matrix.
This means that one innovation can be aligned with more than one vision at a time, for example, if it uses biological processes such as bacteria (biotechnology) to refine a wood based material to replace a fossil input elsewhere in the economy (bioresource vision).
Not all innovations fullfill at least one criteria, yet still originate or are used within the analytical boundaries of the forest based bioeconomy.
I classify these innovations as Vision Neutral.
Examples include innovations aimed at the health and saftey of employees 
<!-- TODO enter source -->
, or general improvements to the longevity of materials and tools used somewhere in the value chain of the forest bioeconomy.
The majority of innovations in this category are of the latter nature.

Innovations were classified as belonging to a bioeconomy vision if they fulfilled one, or more criteria outlined in @bugge2016WhatBioeconomyReview and @vivien2019HijackingBioeconomy.
@tbl-bugge and @tbl-vivien show the summary of bioeconomy visions presented in the cited papers.

**DECISION MATRIX HERE**

For the remainder of the paper the nomenclature of @bugge2016WhatBioeconomyReview is adopted.
Vivien et al.'s [-@vivien2019HijackingBioeconomy] Type I bioeconomy is taken to correspond with the bioecology vision, Type II and Type III to correspond with the bioresource and biotechnology vision, respectively.
The analysis of bioeconomy discourse which revealed different visions, highlights that the visions have emerged at different times and with different actors at their centers [@vivien2019HijackingBioeconomy, @bugge2016WhatBioeconomyReview].
@vivien2019HijackingBioeconomy trace the introduction of the term bioeconomics to the 1920s and a work on fishery by the Russian biologist Baranoff.
The term bioeoconomics which they label as Type I bioeconomy, was introduced by Georgescu-Roegen, in [-@georgescu-roegen1975BioeconomicAspectsEntropy].
@vivien2019HijackingBioeconomy argue that the term biotechnology gradually emerged between 1990 and 2000, in the wake of enthusiasm about biotechnological advances, such as the discovery of DNA.
It gained popularity and international importance when the OECD began increasingly supporting and reshaping its meaning at the turn of the millennium [@vivien2019HijackingBioeconomy].
According to them, the last concept to enter the discourse was the bioresource oriented bioeconomy.
While biomass has been used throughout history in various ways, the renewed interest with the intention to replace fossil inputs, for example in European Commission communication since 2010.
However, @vivien2019HijackingBioeconomy point to the long history of biorefinery as a core aspect of this vision, which was discussed as early as the 1930s.
Hence, while the organization of the discourse into the labels biotechnology, bioecology and bioresource vision is a very recent research outcome, the underlying concepts have a richer history.
This alleviates concerns that using these labels retroactively confines the innovations to current policy terms.

While following the criteria of @bugge2016WhatBioeconomyReview and @vivien2019HijackingBioeconomy, a number of decisions were made concerning content of the summary presented in @tbl-bugge and @tbl-vivien, and the nature of the corresponding vision.
For one, economic growth is cited as the aim and objective of biotechnology and bioresource visions by @bugge2016WhatBioeconomyReview.
It can be argued that all innovation can contribute to economic growth, which is why these aspects were not given heightened importance in the classification decision.
A second topic of judgment has to be the inclusion of innovation which increases the recyclability of a product, or uses recycled waste, water and material in the process.
Based on the definition of a bioecology vision presented in @bugge2016WhatBioeconomyReview, these innovations should be included as contributing to this vision.
The extent to which such innovation contributes in reality to a "[c]ircular and self-sustained production mode" [@bugge2016WhatBioeconomyReview 10], however, can be debated.

@@

Querying the SWINNO database for forest based bioeconomy innovations initally yields `{python} sum_bioeconomy_innovations` innovations. This number includes false positives from substrings matching keywords included in the search. While there is no indication that these false positives are non-randomly distributed across time and sectors, the following analysis for bioeconomy innovation trends at large needs to be viewed in the light of this being unlceaned data. Where the analysis turns to specific bioeconomy visions, only clean and certain cases are discussed.

@@

While the bioeconomy innovation system can be viewed as a niche in the transition to a economic system which does not rely on fossil based technologies, this analysis highlights the layered nature of contested transitions.
Within the bioeconomic innovation system, the different visions could reasonably be conceived of as niches themselves, vying not only for importance over each other, but even for providing sufficient direction to overcome general purpose improvements to and from a regime system.

>>>>

format:
    pdf:
        documentclass: scrartcl
        papersize: A4
        linestretch: 1.25
---

@@

Although the bioecononmy has gained importance as a research, policy and industry focus, it remains unclear what exactly it entails.
Broadly speaking, the bioeconomy refers to an economy in which raw materials for energy and production processes are provided from biological instead of fossil sources.
<!-- TODO: citation -->
Hopes of increased resource efficiency and renewability as well as promised economic growth especially in regions rich in biological resources abound [@el-chichakli2016PolicyFiveCornerstones].
However, a such a profound change to economic systems will not be possible without balancing significant tradeoffs between the needs and desires of different human and non-human stakeholders [@dieken2021MultitudesBioeconomiesSystematic]. @hogbom2021TrilemmaNordicBaltic summarize these tradeoffs as a trilemma between forests ability to provide sufficent wood to replace fossil based materials, carbon sequestration and storage, and the delivery of other eco-system services such as forest biodiversity.
Resolving these tradeoffs is challenging, as goal conflicts between different development pathways make simulatneous progress on social, ecological and economic goals difficult [@bennich2021BiobasedEconomy2030].

@@

These pillars reflect the two most influential visions of the bioeconomy concept. Notions of a biotechnology and bio-resource oriented bioeconomy are widespread, dominate discourse at various levels [@dieken2021MultitudesBioeconomiesSystematic] and form the main industrial visions [@befort2020GoingDefinitionsUnderstand]. 
A third vision, centered around ecology and an economy compatible with the biosphere acts as a counter point.

@@

In more recent years, visions of a bio-resource bioeconomy have taken over as the leading narrative [@allain2022BioeconomicsVsBioeconomy]. Unlike the biotechnology vision, this conceptualization does not only focus on economic growth and job creation, but attempts to include ecological sustainability challenges [@bugge2016WhatBioeconomyReview]. Biological resources, not technologies, form the basis of a new economy. In essence, there are two ways in which bio-resources are poised to replace fossil resources: as substitution, e.g., biofuels replacing fossil fuels, and as high value added products, e.g., cross-laminated timber used as an alternative to concrete in construction [@vivien2019HijackingBioeconomy]. To substitute fossil resources requires significant efforts in innovation. In contrast to the biotechnology vision, however, innovation is not confined to scientific advances from a narrow biochemical and pharmaceutical field. Instead, new and old actors are required to collaborate between and beyond established sectors [@bugge2016WhatBioeconomyReview]. Although sustainability plays a more important role in this vision of the bioeconomy, this vision is considered an example of weak sustainability [@damato2017GreenCircularBio; @vivien2019HijackingBioeconomy]. The reason is the focus of substitution and high biomass demands. The consumption of raw materials may even increase, not only for bio-based, but also for fossil-based resources [@asada2020BioeconomicTransitionProjecting].

@@

For the remainder of the paper the nomenclature of @bugge2016WhatBioeconomyReview is adopted.
Vivien et al.'s [-@vivien2019HijackingBioeconomy] Type I bioeconomy is taken to correspond with the bioecology vision, Type II and Type III to correspond with the bioresource and biotechnology vision, respectively.
The analysis of bioeconomy discourse which revealed different visions, highlights that the visions have emerged at different times and with different actors at their centers [@vivien2019HijackingBioeconomy, @bugge2016WhatBioeconomyReview].
@vivien2019HijackingBioeconomy trace the introduction of the term bioeconomics to the 1920s and a work on fishery by the Russian biologist Baranoff.
The term bioeoconomics which they label as Type I bioeconomy, was introduced by Georgescu-Roegen, in -@georgescu-roegen1975BioeconomicAspectsEntropy.
@vivien2019HijackingBioeconomy argue that the term biotechnology gradually emerged between 1990 and 2000, in the wake of enthusiasm about biotechnological advances, such as the discovery of DNA.
It gained popularity and international importance when the OECD began increasingly supporting and reshaping its meaning at the turn of the millennium [@vivien2019HijackingBioeconomy].
According to them, the last concept to enter the discourse was the bioresource oriented bioeconomy.
While biomass has been used throughout history in various ways, the renewed interest with the intention to replace fossil inputs, for example in European Commission communication since 2010.
However, @vivien2019HijackingBioeconomy point to the long history of biorefinery as a core aspect of this vision, which was discussed as early as the 1930s.
Hence, while the organization of the discourse into the labels biotechnology, bioecology and bioresource vision is a very recent research outcome, the underlying concepts have a richer history.

@@

### Directionality {#sec-directionality}

@@

@weber2012LegitimizingResearchTechnology synthesis policy insights from innovation systems and multilevel perspective research into a failures framework.
Therein, directionality failure is understood as the failure to provide direction to the process of search and unclear or missing setting of collective priorities.
They caution that it should not be confused with shortsighted or flawed anticipation of market developments, as this myopia lacks a view on the direction of change.
To overcome this failure, that is, to provide directionality, they argue, guiding orientation needs to be translated and intermediated on two levels.
External requirements placed on the innovation system need to be absorbed after which they need to be interpreted in a process of negotiation between the different actors.
Reaching this collective understanding of direction, @weber2012LegitimizingResearchTechnology [p. 1042] write, is in practice, "achieved by establishing shared future visions".

@@

Based on the analyses of competing bioeconomy visions in @bugge2016WhatBioeconomyReview and @vivien2019HijackingBioeconomy, I develop and apply a matrix of characterstics for each vision.
 Visions based on @bugge2016WhatBioeconomyReview and @vivien2019HijackingBioeconomy, biotechnology definition based on @yeung2019CurrentResearchBiotechnology. 
<!-- TODO: Matrix That matrix is presented in @tbl- -->
Reading the source texts of innovations identified from SWINNO, I classify an innovation as being *aligned* with a vision if it fulfills at least one criteria in the matrix.
This means that one innovation can be aligned with more than one vision at a time, for example, if it uses biological processes such as bacteria (biotechnology) to refine a wood based material to replace a fossil input elsewhere in the economy (bioresource vision).
Not all innovations fullfill at least one criteria, yet still originate or are used within the analytical boundaries of the forest based bioeconomy.
I classify these innovations as Vision Neutral.
Examples include innovations aimed at the health and saftey of employees 
<!-- TODO enter source -->
, or general improvements to the longevity of materials and tools used somewhere in the value chain of the forest bioeconomy.
The majority of innovations in this category are of the latter nature.

{{< include _decision-matrix.md >}}
: Decision matrix used to classify innovation. {#tbl-decision-matrix tbl-colwidths="[20,80]"}

@@

Querying the SWINNO database for forest based bioeconomy innovations initially yields `{python} sum_bioeconomy_innovations` innovations. This number includes false positives from substrings matching keywords included in the search. While there is no indication that these false positives are non-randomly distributed across time and sectors, the following analysis for bioeconomy innovation trends at large needs to be viewed in the light of this being uncleaned data. Where the analysis turns to specific bioeconomy visions, only clean and certain cases are discussed.

@@

Within the bioeconomic innovation system, the different visions could reasonably be conceived of as niches themselves, vying not only for importance over each other, but even for providing sufficient direction to overcome general purpose improvements to and from a regime system.
While the bioeconomy innovation system can be viewed as a niche in the transition to a economic system which does not rely on fossil based technologies, this analysis highlights the layered nature of contested transitions.
